### 2021-09-22 chapter2/chapter.tex
remove denominator in V(x) operator

%!TEX root = ../thesis.tex
\chapter{A stochastic modelling approach to approximation\label{sec: construction and modelling}}
In this work we develop a new discretisation of a fluid queue using a \emph{quasi-birth-and-death process with rational arrival process components} (QBD-RAP), to be described in more detail later, and prove convergence of the approximation scheme. To improve upon the discretisation of Bean and O'Reilly \cite{bo2013}, we argue that the shift from QBD to QBD-RAP is necessary. The approximation can be made arbitrarily accurate. Due to its stochastic interpretation, the QBD-RAP discretisation ensures solutions are positive, even when discontinuities are present. Unlike slope limiters and post-hoc filtering, for the QBD-RAP the positivity preserving nature is a property of the discretised operator.

This paper details the construction of the approximation and proves its convergence. The QBD-RAP is constructed using \emph{matrix-exponential distributions}. The main result shows the convergence of the approximation scheme under the assumption that the variance of the matrix-exponential distribution(s) used in the construction tends to 0. Given that there is no known simple closed form for the transient distributions of a fluid queue, and that we make minimal assumptions about the matrix-exponential distributions used in the construction of the approximation, it is somewhat remarkable that we are able to establish a convergence result. The generality of the result with respect to the assumptions on the matrix-exponential distributions used in the construction of the approximation is necessitated by the fact that we use a class of \emph{concentrated matrix-exponential distributions} found numerically in \cite{hhat2020}, and for which there is relatively little known about their properties.

The structure of this paper is as follows. In Section~\ref{sec: prelim} we provide a brief background on matrix exponential random variables and the stochastic processes appearing in this work, namely fluid queues, quasi-birth-and-death (QBD) processes, rational arrival processes (RAPs), and quasi-birth-and-death processes with rational arrival process components (QBD-RAPs). Section~\ref{sec: construction and modelling} describes the modelling approach taken to develop the approximation. Section~\ref{sec: inspiration} motivates the idea using the QBD approximation of \cite{bo2014}. Sections \ref{sec: modelling} and \ref{sec: zero states} describe the modelling of certain events of the fluid queue with matrix exponential distributions to ultimately construct the behaviour of the QBD-RAP. Section~\ref{sec: dynamics} describes the dynamics of the resulting construction and constructs the level process of the QBD-RAP. Section~\ref{sec: boundary conditions} deals with modelling boundary behaviour. Section~\ref{sec: initial conditions} describes how to model initial conditions of the fluid queue. Section~\ref{sec: closing} introduces the concept of a \emph{closing operator} which maps states of the QBD-RAP to estimates of the distribution of the fluid queue. We then proceed to show convergence of the approximation scheme in Section~\ref{sec: conv}. First, in Section~\ref{sec: no change}, we derive expressions for certain Laplace transforms with respect to time of the QBD-RAP and fluid queue on the event that there is no change of level. Section~\ref{sec: no change convergence} provides error bounds between the Laplace transforms of the QBD-RAP and the fluid queue and also some domination conditions used so that we may apply the Dominated Convergence Theorem. Section~\ref{sec: local to global} uses the results of the preceding section to prove convergence Laplace transform with respect to time of the expected local time of the QBD-RAP to the Laplace transform with respect to time of the expected local time of the fluid queue. Technical proofs which do not offer a great deal of insight are reserved for appendices.

@@

\subsection{Matrix-exponential distributions}
Here we recount some facts about matrix exponential distributions. See \cite{MEinAP} for a more detailed exposition. A random variable, \(Z\), is said to have a matrix-exponential distribution if it has a distribution function of the form \(1-\bs \alpha e^{\bs Sx}(-\bs S)^{-1}\bs s\), where \(\bs \alpha\) is a \(1\times p\) \emph{initial vector}, \(\bs S\) a \(p\times p\) matrix, and \(\bs s\) a \(p\times 1\) \emph{closing vector}, and \(\displaystyle e^{\bs S x} := \sum_{n=0}^\infty \cfrac{\left(\bs Sx\right)^n}{n!}\) is the matrix exponential. The density function of \(Z\) is given by \(f_Z(x) = \bs\alpha e^{\bs S x}\bs s\). The only restrictions on the parameters \((\bs \alpha, \bs S, \bs s)\) are that \(\bs \alpha e^{\bs S x} \bs s\) be a valid density function, i.e.~\(\bs \alpha e^{\bs S x} \bs s\geq 0,\) for all \(x\geq 0\) and \(\lim_{x\to\infty} 1-\bs \alpha e^{\bs Sx}(-\bs S)^{-1}\bs s = 1.\) There is the possibility of an \emph{atom} (a point mass) at 0, but here we do not consider this possibility. These condition that \(\bs \alpha e^{\bs Sx}\bs s\) be a valid density imposes some properties on representations \((\bs \alpha, \bs S, \bs s)\). However, in general there is no way to determine whether, given a triplet \((\bs \alpha, \bs S, \bs s)\) whether it is a representation of a matrix-exponential distribution, or not. Nonetheless, some properties of a triplet \((\bs \alpha, \bs S, \bs s)\) are known, such as the following, which is used in the characterisation of QBD-RAPs. 
\begin{thm}[Theorem 4.1.3, \cite{MEinAP}]
	The density function of a matrix-exponential distribution with representation \((\bs\alpha, \bs S, \bs s)\) can be expressed in terms of real-valued constants as 
	\begin{align}
		\psi(x)&=\sum_{j=1}^{m_1} \sum_{k=1}^{p_j} c_{jk} \cfrac{x^{k-1}}{(k-1)!} e^{\mu_jx} + \sum_{j=1}^{m+2}\sum_{k=1}^{q_j} d_{jk}\cfrac{x^{k-1}}{(k-1)!}e^{\eta_jx}\cos(\sigma_jx) + \sum_{j=1}^{m_2}\sum_{k=1}^{q_j}e_{jk}\cfrac{x^{k-1}}{(k-1)!}e^{\eta_jx}\sin(\sigma_jx), \label{eqn: spec}
	\end{align}
	for integers \(m_1,\,m_2,\,p_j,\) and \(q_j\) and some real constants \(c_{jk},\,d_{jk},\,e_{jk},\,\mu_j,\,\eta_j,\) and \(\sigma_j\). Here \(\mu_j,\, j=1,\dots,m_1\) are the real eigenvalues of \(\bs S\), while \(\eta_j+i\sigma_j, \, \eta_j-i\sigma_j,\, j=1,\dots,m_2\) denote its complex eigenvalues, which come in conjugate pairs. Thus \(m_1+2m_2\) is the total number of eigenvalues, while the dimension of the representation is given by \(\displaystyle p=\sum_{j=1}^{m+1}p_j + 2\sum_{j=1}^{m_2}q_j\). 
\end{thm}
\begin{thm}[Theorem 4.1.4, \cite{MEinAP}]\label{thm: 4.1.4}
	Consider the nonvanishing terms of the matrix exponential density (\ref{eqn: spec}), \emph{i.e.}, the terms for which \(c_{jk}=0,\,d_{jk} =0,\) or \(e_{jk}=0\). Among the corresponding eigenvalues \(\lambda_j\), there is a real dominating eigenvalue \(\kappa\), say. That is, \(\kappa\) is real, \(\kappa \geq Re(\lambda_j)\) for all \(j\), and the multiplicity of \(\kappa\) is at least the multiplicity of every other eigenvalue with real part \(\kappa\).
\end{thm}
\begin{cor}[Corollary 4.1.5, \cite{MEinAP}]
If \((\bs \alpha, \bs S, \bs s)\) is a representation for a matrix-exponential distribution, then \(\bs S\) has a real dominating eigenvalues.
\end{cor}
\begin{thm}[Theorem 4.1.6, \cite{MEinAP}]
	Let \(Z\) be a matrix-exponentially distributed random variable with density (\ref{eqn: spec}). Then the dominant real eigenvalue \(\kappa\) of Theorem~\ref{thm: 4.1.4} is strictly negative. 
\end{thm}
We define \(dev(\bs S)\) to be the real dominating eigenvalue of \(\bs S\), that is \(dev(\bs S)=\kappa\) in Theorem~\ref{thm: 4.1.4}.

@@

A \emph{representation} of a matrix exponential distribution is a triplet \((\bs \alpha, \bs S, \bs s)\), and we write \(Z\sim ME(\bs \alpha, \bs S, \bs s)\) to denote that \(Z\) has a matrix-exponential distribution with this representation. The order of the representation is the dimension of the square matrix \(\bs S\), i.e.~if \(\bs S\) is \(p\times p\), then the matrix exponential distribution is said to be of order \(p\). Representations of matrix-exponential distributions are not unique \cite{MEinAP}. A representation is called \emph{minimal} when \(\bs S\) has the smallest possible dimension. Throughout this work, we assume that the representation of any matrix exponential distribution is minimal. Let \(\bs e_i\) be a vector with a 1 in the \(i\)th position and zeros elsewhere. We assume that \(\bs s = -\bs S\bs e\), and that \((\bs e_i,\bs S,\bs s)\) for \(i=1,\dots,p\) are representations of matrix exponential distributions. It is always possible to find such a representation \cite[Theorem 4.5.17, Corollary 4.5.18]{MEinAP}. As such, we abbreviate our notation \(Z\sim ME(\bs \alpha, \bs S, \bs s)\) to \(Z\sim ME(\bs \alpha, \bs S)\). Further, given \(\bs s=-\bs S\bs e\) then observe that  \(\displaystyle \int_{x=0}^\infty e^{\bs S x} \bs s\wrt x = (-\bs S)^{-1}\bs s= \bs e\).

@@

Also from Bean and Nielsen \cite{bn2010}, associated with a Marked RAP is a row-vector-valued \emph{orbit} process, \(\{\bs A(t)\}_{t\geq0},\)
\[\bs A(t) = \cfrac{\bs \alpha\left( \prod\limits_{i=1}^{N(t)} e^{\bs S(Y_{i}-Y_{i-1})}\bs D_{M_i}\right)e^{\bs S (t-Y_{N(t)})}}{\bs \alpha\left( \prod\limits_{i=1}^{N(t)} e^{\bs S(Y_{i}-Y_{i-1})}\bs D_{M_i}\right)e^{\bs S( t-Y_{N(t)})}\bs e}.\]
Thus, \(\{\bs A(t)\}\) is a piecewise-deterministic Markov process where, in between events \(\{\bs A(t)\}\) evolves deterministically according to 
\[\bs A(t) = \cfrac{\bs A(Y_{N(t)}^-)e^{\bs S(t-Y_{N(t)})}}{\bs A(Y_{N(t)}^-)e^{\bs S(t-Y_{N(t)})}\bs e},\]
where \(\bs A(Y_{N(t)}^-) = \lim_{u\to 0^+}\bs A(Y_{N(t)}-u)\). The process \(\{\bs A(t)\}\) "jumps" at event times of \(N\) (the process may not always jump at these times, but typically the dynamics change discontinuously at this point). At time \(t\) the intensity with which \(\{\bs A(t)\}\) has a jump is \(\bs A(t) \bs D\bs e\). Upon an event the event is associated with mark \(i\) with probability \(\bs A(t) \bs D_i\bs e/\bs A(t) \bs D\bs e\). Upon on an event at time \(t\) with mark \(i\), the new position of the orbit is \(\bs A(t) = \bs A(t^-) \bs D_i/\bs A(t^-) \bs D_i\bs e\).

@@

\section{Inspiration and motivation}\label{sec: inspiration}
The inspiration for our approach stems from the QBD-based discretisation of Bean and O'Reilly \cite{bo2013}. A QBD can be seen as a two dimensional CTMC, \(\{(L(t),\phi(t))\}_{t\geq 0}\), where \(\{L(t)\}\) is the discrete level, and \(\{\phi(t)\}\) is the phase. Bean and O'Reilly \cite{bo2013} discretise the state space of the fluid into small intervals of width \(\Delta\); denote these by \(\mathcal D_k=[k\Delta,(k+1)\Delta]\). Their approximation captures the dynamics of the phase process exactly, but discretises the level process of the fluid queue. Their approximation supposes that when \(L(t)=k,\,\phi(t)=i\), then \(X(t)\approx k\Delta,\,\phi(t)=i\).

@@

For a given \(\Delta\), the discretisation of Bean and O'Reilly supposes that, when the phase is \(i\) and on the event of no change of phase, the sojourn time in a given level has an exponential distribution with rate \(|c_i|\Delta\). However, the corresponding sojourn time of the fluid queue is deterministic: given \(X(t) = x \in\calD_k\), then \(\{X(t)\}\) will leave \(\calD_k\) in exactly \(((k+1)\Delta - x)/|c_i|\) units of time if \(c_i>0\), and \((x-k\Delta)/|c_i|\) units of time if \(c_i<0\), on the event that the phase does not change before this time. We can extend the QBD model of Bean and O'Reilly to model this determinism more accurately by supposing that the sojourn times in each interval have Erlang distributions (rather than exponential). However, we effectively realise the same QBD approximation as the original, but on a finer discretisation. For example, using Erlang distributions of order 2 we would get a generator of the QBD approximation
\[\bs B = \left[ \begin{array}{cccc} \ddots & \ddots & \ddots & \\ \bs B_{-1}(\Delta) & \bs B_0(\Delta) & \bs B_{+1}(\Delta) & \\  & \bs B_{-1}(\Delta) & \bs B_0(\Delta) & \bs B_{+1}(\Delta) \\ & \ddots & \ddots & \ddots \end{array} \right]\]
\begin{align}
\bs B_0 &= \bs T\otimes I_2 - \left[ \begin{array}{ccc} \left[\begin{array}{cc} -\Delta/2 & \Delta/2 \\ 0 & -\Delta/2 \end{array}\right] \otimes \bs C_+ & & \\ & \left[\begin{array}{cc} -\Delta/2 & 0 \\ \Delta/2 & -\Delta/2 \end{array}\right] \otimes C_- & \\ & & \bs 0 \end{array}\right], 
\end{align}
\begin{align}
\bs B_{-1} &= \left[\begin{array}{ccc} \bs 0 & & \\ & \bs C_-\otimes \left[\begin{array}{cc} 0 & \Delta/2 \\ 0 & 0 \end{array}\right] & \\ & & \bs 0 \end{array}\right],
&&\bs B_{+1} = \left[\begin{array}{ccc} \bs C_+\otimes \left[\begin{array}{cc} 0 & 0 \\ \Delta/2 & 0 \end{array}\right] & & \\ & \bs 0 & \\ & & \bs 0\end{array}\right].
\end{align}
But this is the same discretisation we would get if we were to use intervals of width \(\Delta/2\) in the QBD approximation of Bean and O'Reilly \cite{bo2013}, with the rows/columns reordered. Indeed, this appears to be the best we can do with a QBD-MAP approximation as the Erlang distribution is known to be the least variable Phase-type distribution for a given order \cite{as1987} and therefore gives the closest approximation to the deterministic behaviour we are trying to approximate. This necessitates the extension to QBD-RAPs, whereby we can find more concentrated distributions than any Phase-type, but retain (some of) the stochastic interpretation and matrix-analytic tools.

Here, the construction of the approximating QBD-RAP is developed intuitively, from a stochastic modelling perspective. The key observation is that, given \(\{\varphi(t)\}\) is constant, then \(\{X(t)\}\) moves deterministically. Upon discretising the state space of the level process into intervals of width \(\Delta\), then, given \(X(t)\), the distribution of time it takes for \(\{X(t)\}\) to leave a given interval on the event that \(\{\varphi(t)\}\) remains in the same phase, is deterministic. We model this deterministic behaviour approximately by concentrated matrix exponential distributions (matrix exponential distributions with low-variance). As in Bean and O'Reilly \cite{bo2013}, the discretisation of the state space of \(\{X(t)\}\) corresponds to the level process of a QBD-RAP. Unlike Bean and O'Reilly \cite{bo2013}, however, here we do not take \(\Delta \to 0\). Rather, we suppose that the variance of the concentrated matrix exponential distributions we use to approximate deterministic events gets small. This necessarily means that the order of the matrix exponential increases.

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{chapter2/figs/ME_residual_life_density.pdf}
\caption{The density function for a \emph{concentrated matrix exponential of order 21} from \cite{hhat2020} (blue) and corresponding density functions of the residual lives, \(R_{0.3}\) (red), \(R_{0.5}\) (blue). Observe how the density function of the \(Z_i\) (blue) to approximate a point mass at \(\Delta=1\), while the density functions of \(R_{0.3}\) (red) and \(R_{0.5}\) (blue) approximate point masses at \(0.7\) and \(0.4\), respectively. }
\label{fig:residual distributions}
\end{figure}

@@

Observe that \(\bs D\bs e = \displaystyle \int_{x=0}^\infty e^{\bs S x }\bs s \cfrac{\bs\alpha e^{ \bs Sx } }{\bs\alpha e^{ \bs Sx } \bs e}\wrt x\bs e =\displaystyle \int_{x=0}^\infty e^{\bs S x }\bs s \wrt x=\bs e.\) This is an important observation as it implies that events corresponding to jumps from \(i\in\calS_+\) to \(j\in\calS_-\) occur with intensity \(\bs A(t)T_{ij}\bs D\bs e = T_{ij}\) which is constant with respect to orbit position and time and this implies they are exponential. Further, since \(\mathcal A\) is closed and convex \cite{MEinAP}, then \((\bs a \bs D,\bs S,\bs s)\) is a representation of a matrix exponential distribution for any \(\bs a\in\mathcal A\).

@@

With this construction, and choosing \(Z\sim Erlang(p,\Delta/p)\), we recover the discretisation of Bean and O'Reilly \cite{bo2014} with discretisation parameter \(\Delta/p\).

@@

\section{Computing \(\bs D\)}
In practice we use the class of concentrated matrix exponential distributions (CMEs) found numerically in \cite{hhat2020}. For this class of CMEs, we take the index \(p\) to be the order of the representation. Since the variance of \(Z^{(2p)}\) and \(Z^{(2p-1)}\) are relatively similar \cite{hhat2020}, we take \(p\) to be odd. For a given CME, with representation \((\bs\alpha,\bs S)\), the matrix \(\bs S\) has one real eigenvalue, and \(p-1\) complex eigenvalues and all eigenvalues have the same real part.

@@

Computationally efficient expressions for \(e^{\bs{S}t}\bs e\) and \(\bs a(t)\) are provided in \cite{hhat2020}.

@@

%We approximate the fluid model by the QBD process \(\boldsymbol J(t) = (L(t),\varphi(t))\) with state space \(\mathbb Z\times \mathcal S\) and generator defined by the block matrices \(A_0 = \bs{T}- |\bs{C}|/\Delta\), \(A_1 = |\bs{C}_+|/\Delta\), \(A_{-1} = |\bs{C}_-|\Delta\), where \(|\cdot|\) takes the absolute values of the elements of the matrix. The phase variable of \(\bs J(t)\) and \(\bs M(t)\) match exactly. The process, \(L(t)\Delta\), approximates the fluid height \(X(t)\), and converges as \(\Delta \to 0\) \cite{bo2013}. We also have the approximations 

@@

%Asmussen and Bladt \cite{ab1999} showed that 

>>>>

%!TEX root = ../thesis.tex
\chapter{A stochastic modelling approach to approximation\label{sec: construction and modelling}}
In this chapter we develop a new discretisation of a fluid queue using a \emph{quasi-birth-and-death process with rational arrival process components} (QBD-RAP), to be described in more detail later, and prove convergence of the approximation scheme. To improve upon the discretisation of \cite{bo2013}, we argue that the shift from QBD to QBD-RAP is necessary. The approximation can be made arbitrarily accurate. Due to its stochastic interpretation, the QBD-RAP discretisation ensures solutions are positive, even when discontinuities are present. Unlike slope limiters and post-hoc filtering, for the QBD-RAP the positivity preserving nature is a property of the discretised operator.

The structure of this chapter is as follows. In Section~\ref{sec: prelim} we provide a brief background on matrix exponential random variables and the stochastic processes appearing in this work, namely fluid queues, quasi-birth-and-death (QBD) processes, rational arrival processes (RAPs), and quasi-birth-and-death processes with rational arrival process components (QBD-RAPs). Section~\ref{sec: inspiration} motivates the idea using the QBD approximation of \citep{bo2014}. Sections \ref{sec: modelling} and \ref{sec: zero states} describe the modelling of certain events of the fluid queue with matrix exponential distributions to ultimately construct the behaviour of the QBD-RAP. Section~\ref{sec: dynamics} describes the dynamics of the resulting construction and constructs the level process of the QBD-RAP. Section~\ref{sec: boundary conditions} deals with modelling boundary behaviour. Section~\ref{sec: initial conditions} describes how to model initial conditions of the fluid queue. Section~\ref{sec: closing} introduces the concept of a \emph{closing operator} which maps states of the QBD-RAP to estimates of the distribution of the fluid queue.

@@

\subsection{Matrix-exponential distributions}
Here we recount some facts about matrix exponential distributions. See \citep{MEinAP} for a more detailed exposition. A random variable, \(Z\), is said to have a matrix-exponential distribution if it has a distribution function of the form \(1-\bs \alpha e^{\bs Sx}(-\bs S)^{-1}\bs s\), where \(\bs \alpha\) is a \(1\times p\) \emph{initial vector}, \(\bs S\) a \(p\times p\) matrix, and \(\bs s\) a \(p\times 1\) \emph{closing vector}, and \(\displaystyle e^{\bs S x} := \sum_{n=0}^\infty \cfrac{\left(\bs Sx\right)^n}{n!}\) is the matrix exponential. The density function of \(Z\) is given by \(f_Z(x) = \bs\alpha e^{\bs S x}\bs s\). The only restrictions on the parameters \((\bs \alpha, \bs S, \bs s)\) are that \(\bs \alpha e^{\bs S x} \bs s\) be a valid density function, i.e.~\(\bs \alpha e^{\bs S x} \bs s\geq 0,\) for all \(x\geq 0\) and \(\lim_{x\to\infty} 1-\bs \alpha e^{\bs Sx}(-\bs S)^{-1}\bs s = 1.\) There is the possibility of an \emph{atom} (a point mass) at 0, but here we do not consider this possibility. These condition that \(\bs \alpha e^{\bs Sx}\bs s\) be a valid density imposes some properties on representations \((\bs \alpha, \bs S, \bs s)\). However, in general there is no way to determine whether, given a triplet \((\bs \alpha, \bs S, \bs s)\) whether it is a representation of a matrix-exponential distribution, or not. Nonetheless, some properties of a triplet \((\bs \alpha, \bs S, \bs s)\) are known, such as the following, which is used in the characterisation of QBD-RAPs. 
\begin{thm}[Theorem 4.1.3, \cite{MEinAP}]
	The density function of a matrix-exponential distribution with representation \((\bs\alpha, \bs S, \bs s)\) can be expressed in terms of real-valued constants as 
	\begin{align}
		\psi(x)&=\sum_{j=1}^{m_1} \sum_{k=1}^{p_j} c_{jk} \cfrac{x^{k-1}}{(k-1)!} e^{\mu_jx} + \sum_{j=1}^{m+2}\sum_{k=1}^{q_j} d_{jk}\cfrac{x^{k-1}}{(k-1)!}e^{\eta_jx}\cos(\sigma_jx) + \sum_{j=1}^{m_2}\sum_{k=1}^{q_j}e_{jk}\cfrac{x^{k-1}}{(k-1)!}e^{\eta_jx}\sin(\sigma_jx), \label{eqn: spec}
	\end{align}
	for integers \(m_1,\,m_2,\,p_j,\) and \(q_j\) and some real constants \(c_{jk},\,d_{jk},\,e_{jk},\,\mu_j,\,\eta_j,\) and \(\sigma_j\). Here \(\mu_j,\, j=1,\dots,m_1\) are the real eigenvalues of \(\bs S\), while \(\eta_j+i\sigma_j, \, \eta_j-i\sigma_j,\, j=1,\dots,m_2\) denote its complex eigenvalues, which come in conjugate pairs. Thus \(m_1+2m_2\) is the total number of eigenvalues, while the dimension of the representation is given by \(\displaystyle p=\sum_{j=1}^{m+1}p_j + 2\sum_{j=1}^{m_2}q_j\). 
\end{thm}
\begin{thm}[Theorem 4.1.4, \cite{MEinAP}]\label{thm: 4.1.4}
	Consider the nonvanishing terms of the matrix exponential density (\ref{eqn: spec}), \emph{i.e.}, the terms for which \(c_{jk}=0,\,d_{jk} =0,\) or \(e_{jk}=0\). Among the corresponding eigenvalues \(\lambda_j\), there is a real dominating eigenvalue \(\kappa\), say. That is, \(\kappa\) is real, \(\kappa \geq Re(\lambda_j)\) for all \(j\), and the multiplicity of \(\kappa\) is at least the multiplicity of every other eigenvalue with real part \(\kappa\).
\end{thm}
\begin{cor}[Corollary 4.1.5, \cite{MEinAP}]
If \((\bs \alpha, \bs S, \bs s)\) is a representation for a matrix-exponential distribution, then \(\bs S\) has a real dominating eigenvalues.
\end{cor}
\begin{thm}[Theorem 4.1.6, \cite{MEinAP}]
	Let \(Z\) be a matrix-exponentially distributed random variable with density (\ref{eqn: spec}). Then the dominant real eigenvalue \(\kappa\) of Theorem~\ref{thm: 4.1.4} is strictly negative. 
\end{thm}
We define \(dev(\bs S)\) to be the real dominating eigenvalue of \(\bs S\), that is \(dev(\bs S)=\kappa\) in Theorem~\ref{thm: 4.1.4}.

@@

A \emph{representation} of a matrix exponential distribution is a triplet \((\bs \alpha, \bs S, \bs s)\), and we write \(Z\sim ME(\bs \alpha, \bs S, \bs s)\) to denote that \(Z\) has a matrix-exponential distribution with this representation. The order of the representation is the dimension of the square matrix \(\bs S\), i.e.~if \(\bs S\) is \(p\times p\), then the matrix exponential distribution is said to be of order \(p\). Representations of matrix-exponential distributions are not unique \citep{MEinAP}. A representation is called \emph{minimal} when \(\bs S\) has the smallest possible dimension. Throughout this work, we assume that the representation of any matrix exponential distribution is minimal. Let \(\bs e_i\) be a vector with a 1 in the \(i\)th position and zeros elsewhere. We assume that \(\bs s = -\bs S\bs e\), and that \((\bs e_i,\bs S,\bs s)\) for \(i=1,\dots,p\) are representations of matrix exponential distributions. It is always possible to find such a representation \cite[Theorem 4.5.17, Corollary 4.5.18]{MEinAP}. As such, we abbreviate our notation \(Z\sim ME(\bs \alpha, \bs S, \bs s)\) to \(Z\sim ME(\bs \alpha, \bs S)\). Further, given \(\bs s=-\bs S\bs e\) then observe that  \(\displaystyle \int_{x=0}^\infty e^{\bs S x} \bs s\wrt x = (-\bs S)^{-1}\bs s= \bs e\).

@@

Also from \cite{bn2010}, associated with a Marked RAP is a row-vector-valued \emph{orbit} process, \(\{\bs A(t)\}_{t\geq0},\)
\[\bs A(t) = \cfrac{\bs \alpha\left( \prod\limits_{i=1}^{N(t)} e^{\bs S(Y_{i}-Y_{i-1})}\bs D_{M_i}\right)e^{\bs S (t-Y_{N(t)})}}{\bs \alpha\left( \prod\limits_{i=1}^{N(t)} e^{\bs S(Y_{i}-Y_{i-1})}\bs D_{M_i}\right)e^{\bs S( t-Y_{N(t)})}\bs e}.\]
Thus, \(\{\bs A(t)\}\) is a piecewise-deterministic Markov process where, in between events \(\{\bs A(t)\}\) evolves deterministically according to 
\[\bs A(t) = \cfrac{\bs A(Y_{N(t)}^-)e^{\bs S(t-Y_{N(t)})}}{\bs A(Y_{N(t)}^-)e^{\bs S(t-Y_{N(t)})}\bs e},\]
where \(\bs A(Y_{N(t)}^-) = \lim_{u\to 0^+}\bs A(Y_{N(t)}-u)\). The process \(\{\bs A(t)\}\) "jumps" at event times of \(N\) (the process may not always jump at these times, but typically the dynamics change discontinuously at this point). At time \(t\) the intensity with which \(\{\bs A(t)\}\) has a jump is \(\bs A(t) \bs D\bs e\). Upon an event the event is associated with mark \(i\) with probability \(\bs A(t) \bs D_i\bs e/\bs A(t) \bs D\bs e\). Upon on an event at time \(t\) with mark \(i\), the new position of the orbit is \(\bs A(t) = \bs A(t^-) \bs D_i/\bs A(t^-) \bs D_i\bs e\).

@@

\section{Inspiration and motivation}\label{sec: inspiration}
The inspiration for our approach stems from the QBD-based discretisation of \cite{bo2013}. A QBD can be seen as a two dimensional CTMC, \(\{(L(t),\phi(t))\}_{t\geq 0}\), where \(\{L(t)\}\) is the discrete level, and \(\{\phi(t)\}\) is the phase. \cite{bo2013} discretise the state space of the fluid into small intervals of width \(\Delta\); denote these by \(\mathcal D_k=[k\Delta,(k+1)\Delta]\). Their approximation captures the dynamics of the phase process exactly, but discretises the level process of the fluid queue. Their approximation supposes that when \(L(t)=k,\,\phi(t)=i\), then \(X(t)\approx k\Delta,\,\phi(t)=i\).

@@

For a given \(\Delta\), the discretisation of Bean and O'Reilly supposes that, when the phase is \(i\) and on the event of no change of phase, the sojourn time in a given level has an exponential distribution with rate \(|c_i|\Delta\). However, the corresponding sojourn time of the fluid queue is deterministic: given \(X(t) = x \in\calD_k\), then \(\{X(t)\}\) will leave \(\calD_k\) in exactly \(((k+1)\Delta - x)/|c_i|\) units of time if \(c_i>0\), and \((x-k\Delta)/|c_i|\) units of time if \(c_i<0\), on the event that the phase does not change before this time. We can extend the QBD model of Bean and O'Reilly to model this determinism more accurately by supposing that the sojourn times in each interval have Erlang distributions (rather than exponential). However, we effectively realise the same QBD approximation as the original, but on a finer discretisation. For example, using Erlang distributions of order 2 we would get a generator of the QBD approximation
\[\bs B = \left[ \begin{array}{cccc} \ddots & \ddots & \ddots & \\ \bs B_{-1}(\Delta) & \bs B_0(\Delta) & \bs B_{+1}(\Delta) & \\  & \bs B_{-1}(\Delta) & \bs B_0(\Delta) & \bs B_{+1}(\Delta) \\ & \ddots & \ddots & \ddots \end{array} \right]\]
\begin{align}
\bs B_0 &= \bs T\otimes I_2 - \left[ \begin{array}{ccc} \left[\begin{array}{cc} -\Delta/2 & \Delta/2 \\ 0 & -\Delta/2 \end{array}\right] \otimes \bs C_+ & & \\ & \left[\begin{array}{cc} -\Delta/2 & 0 \\ \Delta/2 & -\Delta/2 \end{array}\right] \otimes C_- & \\ & & \bs 0 \end{array}\right], 
\end{align}
\begin{align}
\bs B_{-1} &= \left[\begin{array}{ccc} \bs 0 & & \\ & \bs C_-\otimes \left[\begin{array}{cc} 0 & \Delta/2 \\ 0 & 0 \end{array}\right] & \\ & & \bs 0 \end{array}\right],
&&\bs B_{+1} = \left[\begin{array}{ccc} \bs C_+\otimes \left[\begin{array}{cc} 0 & 0 \\ \Delta/2 & 0 \end{array}\right] & & \\ & \bs 0 & \\ & & \bs 0\end{array}\right].
\end{align}
But this is the same discretisation we would get if we were to use intervals of width \(\Delta/2\) in the QBD approximation of \cite{bo2013}, with the rows/columns reordered. Indeed, this appears to be the best we can do with a QBD-MAP approximation as the Erlang distribution is known to be the least variable Phase-type distribution for a given order \citep{as1987} and therefore gives the closest approximation to the deterministic behaviour we are trying to approximate. This necessitates the extension to QBD-RAPs, whereby we can find more concentrated distributions than any Phase-type, but retain (some of) the stochastic interpretation and matrix-analytic tools.

Here, the construction of the approximating QBD-RAP is developed intuitively, from a stochastic modelling perspective. The key observation is that, given \(\{\varphi(t)\}\) is constant, then \(\{X(t)\}\) moves deterministically. Upon discretising the state space of the level process into intervals of width \(\Delta\), then, given \(X(t)\), the distribution of time it takes for \(\{X(t)\}\) to leave a given interval on the event that \(\{\varphi(t)\}\) remains in the same phase, is deterministic. We model this deterministic behaviour approximately by concentrated matrix exponential distributions (matrix exponential distributions with low-variance). As in \cite{bo2013}, the discretisation of the state space of \(\{X(t)\}\) corresponds to the level process of a QBD-RAP. Unlike  \cite{bo2013}, however, here we do not take \(\Delta \to 0\). Rather, we suppose that the variance of the concentrated matrix exponential distributions we use to approximate deterministic events gets small. This necessarily means that the order of the matrix exponential increases.

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{chapter2/figs/ME_residual_life_density.pdf}
\caption{The density function for a \emph{concentrated matrix exponential of order 21} from \citep{hhat2020} (blue) and corresponding density functions of the residual lives, \(R_{0.3}\) (red), \(R_{0.5}\) (blue). Observe how the density function of the \(Z_i\) (blue) to approximate a point mass at \(\Delta=1\), while the density functions of \(R_{0.3}\) (red) and \(R_{0.5}\) (blue) approximate point masses at \(0.7\) and \(0.4\), respectively. }
\label{fig:residual distributions}
\end{figure}

@@

Observe that \(\bs D\bs e = \displaystyle \int_{x=0}^\infty e^{\bs S x }\bs s \cfrac{\bs\alpha e^{ \bs Sx } }{\bs\alpha e^{ \bs Sx } \bs e}\wrt x\bs e =\displaystyle \int_{x=0}^\infty e^{\bs S x }\bs s \wrt x=\bs e.\) This is an important observation as it implies that events corresponding to jumps from \(i\in\calS_+\) to \(j\in\calS_-\) occur with intensity \(\bs A(t)T_{ij}\bs D\bs e = T_{ij}\) which is constant with respect to orbit position and time and this implies they are exponential. Further, since \(\mathcal A\) is closed and convex \citep{MEinAP}, then \((\bs a \bs D,\bs S,\bs s)\) is a representation of a matrix exponential distribution for any \(\bs a\in\mathcal A\).

@@

With this construction, and choosing \(Z\sim Erlang(p,\Delta/p)\), we recover the discretisation of \cite{bo2014} with discretisation parameter \(\Delta/p\).

@@

\section{Computing \(\bs D\)}
In practice we use the class of concentrated matrix exponential distributions (CMEs) found numerically in \citep{hhat2020}. For this class of CMEs, we take the index \(p\) to be the order of the representation. Since the variance of \(Z^{(2p)}\) and \(Z^{(2p-1)}\) are relatively similar \citep{hhat2020}, we take \(p\) to be odd. For a given CME, with representation \((\bs\alpha,\bs S)\), the matrix \(\bs S\) has one real eigenvalue, and \(p-1\) complex eigenvalues and all eigenvalues have the same real part.

@@

Computationally efficient expressions for \(e^{\bs{S}t}\bs e\) and \(\bs a(t)\) are provided in \citep{hhat2020}.

@@

%We approximate the fluid model by the QBD process \(\boldsymbol J(t) = (L(t),\varphi(t))\) with state space \(\mathbb Z\times \mathcal S\) and generator defined by the block matrices \(A_0 = \bs{T}- |\bs{C}|/\Delta\), \(A_1 = |\bs{C}_+|/\Delta\), \(A_{-1} = |\bs{C}_-|\Delta\), where \(|\cdot|\) takes the absolute values of the elements of the matrix. The phase variable of \(\bs J(t)\) and \(\bs M(t)\) match exactly. The process, \(L(t)\Delta\), approximates the fluid height \(X(t)\), and converges as \(\Delta \to 0\) \citep{bo2013}. We also have the approximations 

@@

%Asmussen and Bladt \citep{ab1999} showed that 
